---
title: "Archvale Data Analysis"
author: "Nianyu Li"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
  word_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

```{r}
#| label: load-libraries
#| echo: false # This option disables the printing of code (only output is displayed).
#| message: false
#| warning: false

library(tidyverse)
library(ggplot2)
library(skimr)
library(here)
library(lubridate)
library(janitor)
library(openxlsx)
library(waterfalls)
library(GGally)
```

# Loading the data

```{r}
#| echo: false
#| message: false
#| warning: false

patient <- read_csv(here::here("data", "patientRecord.csv")) %>% 
  clean_names()
glimpse(patient)

appointment <- read_csv(here::here("data", "appointmentRecord.csv")) %>% 
  clean_names()
glimpse(appointment)

chronicDisease <- read_csv(here::here("data", "chronicDiseaseRecord.csv")) %>% 
  clean_names()
glimpse(chronicDisease)
```

# Data Preparation

## Frequencies

Sum the number of consultations of each patients and join the table with
`patient` table.

```{r}
#Count visit times of each patients
visit_n <- 
  appointment %>% 
  group_by(patient_id) %>%
  summarise(visit_n = n())

#Join the table with patients record
patients_n_visits <- 
  left_join(x = patient,
            y = visit_n,
            by = "patient_id") %>% 
  
  #change NA to 0 for the patients who have never visited GP
  mutate(visit_n = replace_na(visit_n,0)) %>% 
  
  arrange(desc(visit_n)) %>% 
  
  print()
```

## Age

The age of the patient might affect their behaviour. As the time frame
of the appoint record is not that long (only includes record of
2022-2023), we could assume the patient's age is this year minus the
year of their birthday, though not 100% accurate.

```{r}
patients_n_visits_1 <-  patients_n_visits %>% 
  
  #convert character to date
  mutate(date_of_birth = dmy(date_of_birth)) %>% 
  
  #calculate age
  mutate(age = year(today()) - year(date_of_birth)) %>% 
  
  print()
```

```{r}
#Assign patients to "Early_childhood(<8) and elderly(>65)" and "Others"

#Create a new variable to label patients
patients_n_visits_1 <-patients_n_visits_1 %>% 
  mutate(age_group = case_when(age %in% (8:65) ~ "Others",
                               !age %in% (8:65) ~ "Under 8 or above 65")) %>% 
  print()
```

## Registration Cohort

Patients who registered after March 2023 might have a different
behaviour due to the implementation of a marketing campign. Create a new
variable to show the registration cohort of the patients: people who
registered before March 2023 are classified as in the `Old` cohort,
whereas after March 2023 belongs to `New` cohort.

```{r}
patients_n_visits_2 <-  patients_n_visits_1 %>% 
  
  #convert character to date
  mutate(registration_date = dmy(registration_date)) %>%
  
  #create a new variable to show their registration cohort
  mutate(cohort = case_when(registration_date > "2023-03-01" ~ "New",
                            registration_date < "2023-03-01" ~ "Old")) %>% 
  
  print()
```

## Chronic Disease

Patients with chronic diseases visits GP regularly. Create a new
variable called chronic_disease to show the number of disease the
patient have.

```{r}
#convert the data to long format
chronicDisease_1 <- chronicDisease %>% 
  pivot_longer(cols = -"patient_id",
               names_to = "disease",
               values_to = "yn") %>%
  
  group_by(patient_id) %>% 
  summarise(n_disease = sum(yn)) 

patients_n_visits_3 <-
  
  #Join the chronic disease table with patients table
  left_join(x = patients_n_visits_2,
            y = chronicDisease_1,
            by = "patient_id") %>% 
  
  #convert NA to 0
  mutate(n_disease = replace_na(n_disease,0)) %>% 
  
  #label patients as have(Yes) and do not have(No) chronic disease
  mutate(chronic_disease = case_when(n_disease > 0 ~ "Yes",
                                     n_disease == 0 ~ "No")) %>% 
  
  print()
```

## Average appointment time

```{r}
#| warning: false

#Calculate the average appointment time for patients
app_time <- appointment %>% 
  group_by(patient_id) %>% 
  summarise(avg_time = mean(as.integer(planned_appointment_duration)))

#join it with the patients table
patients_n_visits_4 <- left_join(x = patients_n_visits_3,
                                 y = app_time,
                                 by = "patient_id") %>% 
  
  #convert NA to 0
  mutate(avg_time = replace_na(avg_time,0)) %>% 
  
  print()
```

## Average appointment intervals

```{r}
app_interval <- appointment %>% 
  
  #convert character to date
  mutate(appointment_date = dmy(appointment_date)) %>%
  
  #group by patients
  group_by(patient_id) %>% 
  
  #calculate average interval
  summarise(app_interval_sum = max(appointment_date) - min(appointment_date),
            visit_n = n()) %>% 
  
  mutate(app_interval = app_interval_sum/visit_n,
         app_interval_int = as.numeric(app_interval)) %>% 
  
  #arrange the data
  arrange(desc(app_interval)) %>%
  
  select(patient_id, app_interval_int)
  
  #Join it with patients table and omit NAs
patients_n_visits_5 <- left_join(x = patients_n_visits_4,
                                 y = app_interval,
                                 by = "patient_id") %>% 
  
  #convert NA to 0
  mutate(app_interval_int = replace_na(app_interval_int,0)) %>% 
  
  print()
```

# Data Analysis

We make the following assumptions to identify the inappropriate users of
GP services:

1)  They are frequent users of GP services.

2)  They are associate with potential inappropriate usage - assumptions
    about inappropriate usage.

## Identify Frequent Patients

Calculate the patient's GP frequencies to identify the frequent
patients.

```{r}
patients_frequency <- patients_n_visits_5 %>%
  
  #group the patients by their frequencies
  group_by(visit_n) %>% 
  
  #count the total number of patients and visits in each group
  summarise(patient_n = n(),
            visit_total_n = sum(visit_n)) %>% 
  
  #calculate the percentages
  mutate(patient_pct = round(patient_n/sum(patient_n)*100,2),
         visit_pct = round(visit_total_n/sum(visit_total_n)*100,2)) %>% 
  
  #calculate the cumulative percentages
  arrange(desc(visit_n)) %>% 
  mutate(patient_cumpct = cumsum(patient_pct),
         visit_cumpct = cumsum(visit_pct)) %>% 
  
  print()

#save the tibble as an excel
#write.xlsx(patients_frequency, "patients_frequency.xlsx")
```

From the cumulative percentages, we can tell that patients who visit the
GP more than 7 times account for only about 10% of total patients, but
more than 45% of the consultations. Therefore, we name this group as
`Frequent` and the other as `Non-frequent`.

```{r}
patients_frequency %>%

  #summarize percentage of visits of frequent and non-frequent patients

  mutate(patient_type = case_when(visit_n >= 7 ~ "Frequent(>=7)",
                                  visit_n < 7 ~ "Non-frequent(<7)")) %>%
  
  group_by(patient_type) %>%

  summarise(pct_visit_total = sum(visit_pct),
            pct_patient_total = sum(patient_pct)) %>%
  
  ggplot()+

  aes(x = '',
      y = pct_visit_total,
      fill = patient_type) +

  geom_col(position = "stack") +

  scale_fill_brewer()+

  geom_text(aes(label = paste0(pct_visit_total,"%")),
            position = position_stack(vjust = 0.5)) +

  coord_polar("y")+

  theme_void()+

  theme(axis.ticks = element_blank(),
        axis.text = element_blank())+

  labs(title = "% of Consultation by Patient Type",
       x = NULL, y = NULL,
       fill = "Patient Type")

```

Label the patients as `Frequent` and `Non-frequent`.

```{r}
patients_n_visits_6 <- 
  patients_n_visits_5 %>% 
  mutate(segment = case_when(visit_n >= 7 ~ "Frequent",
                             visit_n < 7 ~ "Non-frequent")) %>% 
  print()
```

## Hypothesis and Assumptions

Now that we have frequent users of GP, to find out inappropriate usage,
we need to make a few assumptions about which group in the frequent
users is more likely to use GP inappropriately. In other words, how
likely is frequent usage associate with inappropriate usage for a
particular group of patients.

According to the hypothesis provided, Age/Gender/Chronic
Disease/Registration Cohort might have an impact on the patient's
frequency. We might take that into consideration when we decide whether
the patient has an appropriate clinical need.

### Age: people under 8 or above 65 are not likely to use GP inappropriately

People under 8 or above 65 account for about 1/3 of the Frequent
patients, but only 1/5 of the Non-frequent patient.

```{r}
patients_n_visits_6 %>%
  
  count(age_group = factor(age_group), segment = factor(segment)) %>% 
    mutate(pct = prop.table(n)) %>% 
  
  ggplot()+
  
  aes(x = segment,
      y = pct,
      fill = age_group,
      label = scales::percent(pct))+
  
  geom_col(position = "dodge")+
  
  geom_text(color = "Black",
            position = position_dodge(width = .9), 
            vjust = -0.2, size = 4) + 
  
  scale_y_continuous(labels = scales::percent)+
  
  scale_fill_manual(values = c("#001e62","#c6cddc"))+
  
  theme_minimal()+
  
  theme(legend.position = "bottom")+
  
  labs(title = "Patients frequencies in different age groups",
       x = "Patient Segment", y = NULL,
       fill = "Age Group")
```

### Gender: female are not likely to use GP inappropriately

The bar plot below shows that Female accounts for more frequent
patients.

```{r}
patients_n_visits_6 %>%
  
  count(gender = factor(gender), segment = factor(segment)) %>% 
    mutate(pct = prop.table(n)) %>% 
  
  ggplot()+
  
  aes(x = segment,
      y = pct,
      fill = gender,
      label = scales::percent(pct))+
  
  geom_col(position = "dodge")+
  
  geom_text(color = "Black",
            position = position_dodge(width = .9), 
            vjust = -0.2, size = 4) + 
  
  scale_y_continuous(labels = scales::percent)+
  
  scale_fill_manual(values = c("#001e62","#c6cddc"))+
  
  theme_minimal()+
  
  theme(legend.position = "bottom")+
  
  labs(title = "Patients frequencies in different gender",
       x = "Patient Segment", y = NULL,
       fill = "Gender")
```

### Chronic Disease: people with chronic diseases are not likely to use GP inappropriately

People with chronic diseases have a higher proportion of Frequent
patients.

```{r}
patients_n_visits_6 %>%
  
  #first plot for patients who do/does not have chronic disease
  
  count(chronic_disease = factor(chronic_disease), segment = factor(segment)) %>% 
    mutate(pct = prop.table(n)) %>% 
  
  ggplot()+
  
  aes(x = segment,
      y = pct,
      fill = chronic_disease,
      label = scales::percent(pct))+
  
  geom_col(position = "dodge")+
  
  geom_text(color = "Black",
            position = position_dodge(width = .9), 
            vjust = -0.2, size = 4)+ 
  
  scale_y_continuous(labels = scales::percent)+
  
  scale_fill_manual(values = c("#001e62","#c6cddc"))+
  
  theme_minimal()+
  
  theme(legend.position = "bottom")+
  
  labs(title = "Patients frequencies by whether they have chronic disease",
       x = "Patient Segment", y = NULL,
       fill = "Whether the patient has chronic disease")
```

The more chronic diseases the patient has, the less likely it will be a
Non-frequent patient.

```{r}
patients_n_visits_6 %>%
  
  #then only plot for patients who have chronic disease
  filter(chronic_disease == "Yes") %>% 
  
  ggplot()+
  
  aes(x = n_disease,
      fill = segment)+
  
  geom_bar(stat = "count")+
  
  scale_fill_manual(values = c("#001e62","#c6cddc"))+
  
  scale_x_continuous(breaks = c(1:7))+
  
  theme_minimal()+
  
  theme(legend.position = "bottom")+
  
  labs(title = "Patients frequencies by # of chronic disease they have",
       x = "How many chronic disease the patient has", y = NULL,
       fill = "Segment")
```

### Registration Cohort: 10%-50% rule does not apply to the new cohort

```{r}
#Filter data for just people in the "New" cohort
patients_frequency_cohort <- patients_n_visits_6 %>%
  
  filter(cohort == "New") %>% 
  
  #group the patients by their frequencies
  group_by(visit_n) %>% 
  
  #count the total number of patients and visits in each group
  summarise(patient_n = n(),
            visit_total_n = sum(visit_n)) %>% 
  
  #calculate the percentages
  mutate(patient_pct = round(patient_n/sum(patient_n)*100,2),
         visit_pct = round(visit_total_n/sum(visit_total_n)*100,2)) %>% 
  
  #calculate the cumulative percentages
  arrange(desc(visit_n)) %>% 
  mutate(patient_cumpct = cumsum(patient_pct),
         visit_cumpct = cumsum(visit_pct)) %>% 
  
  print()
```

```{r}
#Filter data for just people in the "Old" cohort
patients_frequency_cohort_old <- patients_n_visits_6 %>%
  
  filter(cohort == "Old") %>% 
  
  #group the patients by their frequencies
  group_by(visit_n) %>% 
  
  #count the total number of patients and visits in each group
  summarise(patient_n = n(),
            visit_total_n = sum(visit_n)) %>% 
  
  #calculate the percentages
  mutate(patient_pct = round(patient_n/sum(patient_n)*100,2),
         visit_pct = round(visit_total_n/sum(visit_total_n)*100,2)) %>% 
  
  #calculate the cumulative percentages
  arrange(desc(visit_n)) %>% 
  mutate(patient_cumpct = cumsum(patient_pct),
         visit_cumpct = cumsum(visit_pct)) %>% 
  
  print()
```

As we can see, new cohort does not appears to behave as the 10%-50%
pattern we discussed before, and 93% of the patients haven't visited GP,
therefore, we only filter for old cohort to find inappropriate usage.

### Average appointment time and appointment intervals

I also assume that people with shorter average appointment time and
intervals between appointments are more likely to associate with
inappropriate usage.

## Identify Inappropriate Usage

Filter for frequent patients in the old cohort.
```{r}
patients_oldco <- patients_n_visits_6 %>% 
  
  #Filter only for old cohort
  filter(cohort == "Old"&
           segment == "Frequent") %>% 
  
  print()
```

Filter out the patients using the following features based on the assumptions we discussed above.

    Features include:

-   People who are not "Under 8 or above 65";
-   People who do not have chronic disease;
-   Shorter average planned time;
-   Shorter Intervals between appointments;

```{r}
# first for vulnerable patients
patients_oldco_sus <- patients_oldco %>%
  
  # Shorter average planned time
  filter(avg_time<median(avg_time) | 
         # Shorter Intervals between appointments
         app_interval_int<median(app_interval_int)) %>% 
  
  print()
```

```{r}
patients_oldco_sus_notvul <- patients_oldco_sus %>% 
  filter(age_group == "Others" &
         chronic_disease == "No") %>% 
  print()
```

```{r}
#save the tibble as an excel
#write.xlsx(patients_oldco_sus_notvul, "final_patient_list_1.xlsx")
```

# Reflections

1. Gender does affect patients' frequency, but it was not taken into consideration when assumptions are established.

2. Assumptions are made based on hypothesis provided, plus my understanding of inappropriate behaviour. In usual practices, qualitative analysis is done in the first place to make sure that we fully understand the reasons behind inappropriate usage.

3. I filtered the data to get the final list due to time constraints. Usually, a score card is designed to assess each patients' behaviour, to see how close each case is to our ideal profile - in this case, a potential inappropriate user. Then select the top 50 in the ranking to get the final list.

